\section{Conclusion}
\label{sec:conclusion}

Agentic workflows are becoming programs. They branch, maintain 
state, call tools, delegate subtasks, and increasingly generate 
other workflows. \pact{} makes this observation explicit by 
providing a typed imperative DSL in which \texttt{ask}, 
\texttt{say}, \texttt{gen}, and \texttt{agent} are first-class 
typed expressions rather than unstructured library calls. The 
language is designed around the fact that LLMs do not merely 
produce values. They can also produce program structure, including 
schemas that govern future data, workflows that are later executed, 
and policies that restrict subsequent operations.

The central mechanism is \emph{gradual validation}. A generated 
\texttt{Schema}, arrow-typed workflow, or \texttt{Policy} is 
accepted only provisionally until it is put to use. At that point, 
\pact{} re-checks the dependent program region under the same 
typing and authorization discipline as programmer-written code. 
When validation fails, the runtime retries the causal generator 
with the violated constraint as feedback, rather than aborting the 
workflow or repairing only a downstream symptom. This makes 
validation serve three roles at once. It enforces structural 
correctness of generated artifacts, it localizes recovery for 
long-running workflows, and it avoids repeated dependent failures 
that would otherwise waste tokens.

\pact{} also integrates authorization into the workflow language 
itself. Generated Cedar policies are activated explicitly through 
\texttt{enforce}, installed append-only, and checked before every 
\texttt{gen} and \texttt{agent} invocation under the current 
principal. Thus generated workflows and generated policies remain 
subject to language-level control rather than escaping into an 
untyped orchestration layer. We formalized the resulting runtime 
as a small-step semantics with staged materialization, fail-closed 
behavior, and policy monotonicity, and mechanized the per-step 
soundness argument in Lean~4. A Rust prototype and clinical-trial 
case study show that the same mechanism can be implemented 
directly and can reduce manual checking, runtime failures, and 
repeated downstream retries compared with assertion-based 
baselines.