\section{Design}
\label{sec:design}

This section formalizes the HADL type system. We present the abstract syntax, the type language and subtyping relation, the typing rules, and the staged gradual type checking mechanism that connects the static type system to runtime schema resolution.

\subsection{Syntax}
\label{sec:design-syntax}

Figure~\ref{fig:syntax} presents the abstract syntax of HADL. A program $P$ consists of a sequence of pragmas $\pi$ followed by a sequence of declarations $d$. Pragmas configure runtime behavior without affecting the type system. A declaration is either a workflow declaration, which introduces a named, typed function with parameters and a body expression, or a type declaration, which introduces a named type alias. Types include the unit type $()$, three base types (\texttt{String}, \texttt{Size}, \texttt{Bool}), the gradually-typed \texttt{Schema}, curried arrow types $(T_1 \to T_2)$, and named references $X$ to type declarations or schema-typed variables. Expressions include literals, variable references, immutable and mutable bindings (with optional type annotations), assignment, control flow (\texttt{if}, \texttt{while}, \texttt{for}), the three domain-specific primitives (\texttt{user}, \texttt{tool}, \texttt{agent}), JavaScript interop (\texttt{js}), workflow invocation (\texttt{eval}), comparison operators, and sequencing.

\begin{figure}[t]
\[\begin{array}{rcll}
P & ::= & \pi^*\; d^* & \textit{program} \\[4pt]
\pi & ::= & \texttt{@policy}\; s & \textit{policy pragma} \\
    & \mid & \texttt{@retries}\; n & \textit{retry pragma} \\[4pt]
d & ::= & \texttt{workflow}\; f(\overline{x_i{:}T_i}) \to T\; \{e\} & \textit{workflow declaration} \\
  & \mid & \texttt{type}\; X = T & \textit{type declaration} \\[4pt]
T & ::= & () \mid \texttt{String} \mid \texttt{Size} \mid \texttt{Bool} & \textit{base types} \\
  & \mid & \texttt{Schema} & \textit{gradually-typed unknown} \\
  & \mid & (T_1 \to T_2) & \textit{arrow type} \\
  & \mid & X & \textit{named type reference} \\[4pt]
e & ::= & s \mid n \mid b & \textit{literals} \\
  & \mid & x & \textit{variable reference} \\
  & \mid & \texttt{let}\; x {:} T = e \mid \texttt{let}\; x = e & \textit{immutable binding} \\
  & \mid & \texttt{var}\; x {:} T = e \mid \texttt{var}\; x = e & \textit{mutable binding} \\
  & \mid & x = e & \textit{assignment} \\
  & \mid & \texttt{if}\; (e)\; \{e\}\; [\texttt{else}\; \{e\}] & \textit{conditional} \\
  & \mid & \texttt{while}\; (e)\; \{e\} & \textit{loop} \\
  & \mid & \texttt{for}\; x\; \texttt{in}\; e\; \{e\} & \textit{iteration} \\
  & \mid & \texttt{user}\; s & \textit{human prompt} \\
  & \mid & \texttt{tool}\; s\; [\texttt{as}\; s] & \textit{LLM tool call} \\
  & \mid & \texttt{agent}\; s\; [\texttt{as}\; s] & \textit{LLM agent call} \\
  & \mid & \texttt{js}\; c & \textit{JavaScript expression} \\
  & \mid & \texttt{eval}\; e\;(e_1, \ldots, e_k) & \textit{workflow invocation} \\
  & \mid & e_1 \;\mathit{op}\; e_2 & \textit{comparison} \\
  & \mid & e_1 ;\; e_2 & \textit{sequencing} \\[4pt]
\mathit{op} & ::= & \texttt{==} \mid \texttt{!=} \mid \texttt{<} \mid \texttt{<=} \mid \texttt{>} \mid \texttt{>=} & \textit{comparison operators}
\end{array}\]
\caption{Abstract syntax of HADL. Square brackets denote optional components. Overlined parameters $\overline{x_i{:}T_i}$ denote a possibly empty sequence. Metavariables $s$, $n$, $b$, $c$, $x$, $f$, $X$ range over string literals, size literals, boolean literals, JavaScript code, variable names, workflow names, and type names respectively.}
\label{fig:syntax}
\end{figure}


\subsection{Types and Subtyping}
\label{sec:design-subtyping}

The type language consists of four base types, the gradually-typed \texttt{Schema}, arrow types, and named type references. The unit type $()$ represents expressions evaluated for side effects. \texttt{String} is the type of text values, including all outputs from \texttt{user} and \texttt{agent} expressions. \texttt{Size} is the type of non-negative integers. \texttt{Bool} is the type of boolean values and comparison results. \texttt{Schema} is the gradually-typed unknown, representing values whose concrete type will be determined at runtime by an LLM tool call. Arrow types $(T_1 \to T_2)$ represent workflow function types and are constructed by currying the parameter list of a workflow declaration. Named types $X$ refer to either a \texttt{type} declaration or a schema-typed variable whose concrete type has been registered at runtime.

\paragraph{Type resolution.} Before subtyping or checking, named types are resolved through the type environment. A named type $X$ is first looked up in the type declarations. If found, the result is recursively resolved. If $X$ is not a type declaration but is a variable with type \texttt{Schema} in the environment, resolution returns \texttt{Schema}, reflecting that the concrete type of this variable is deferred to runtime. Arrow types are resolved recursively in both components. If $X$ is neither declared nor a schema-typed variable, a type error is reported.

\paragraph{Subtyping.} Figure~\ref{fig:subtyping} defines the subtyping relation $T_1 \leq T_2$. The relation is reflexive on each base type (\textsc{S-Unit} through \textsc{S-Schema}). The key rules for gradual typing are \textsc{S-Schema-L} and \textsc{S-Schema-R}, which make \texttt{Schema} compatible with every type in both directions. This allows schema-typed values to flow to any position at compile time, deferring the actual type check to runtime when the schema is resolved. Arrow subtyping (\textsc{S-Arrow}) follows standard variance conventions, contravariant in the argument and covariant in the return type.

\begin{figure}[t]
\begin{mathpar}
\inferrule*[right=S-Unit]{ }{() \leq ()}
\and
\inferrule*[right=S-String]{ }{\texttt{String} \leq \texttt{String}}
\and
\inferrule*[right=S-Size]{ }{\texttt{Size} \leq \texttt{Size}}
\and
\inferrule*[right=S-Bool]{ }{\texttt{Bool} \leq \texttt{Bool}}
\and
\inferrule*[right=S-Schema]{ }{\texttt{Schema} \leq \texttt{Schema}}
\and
\inferrule*[right=S-Schema-L]{ }{\texttt{Schema} \leq T}
\and
\inferrule*[right=S-Schema-R]{ }{T \leq \texttt{Schema}}
\and
\inferrule*[right=S-Arrow]{A_2 \leq A_1 \\ R_1 \leq R_2}{(A_1 \to R_1) \leq (A_2 \to R_2)}
\end{mathpar}
\caption{Subtyping rules. \textsc{S-Schema-L} and \textsc{S-Schema-R} make \texttt{Schema} the gradually-typed unknown, compatible with all types. Concrete type checking is deferred to runtime when the schema value is resolved.}
\label{fig:subtyping}
\end{figure}


\subsection{Typing Rules}
\label{sec:design-typing}

The type checker operates in two modes, expressed as two judgments. The \emph{synthesis} judgment $\Gamma \vdash e \Rightarrow T$ takes an environment and an expression and produces a type. The \emph{checking} judgment $\Gamma \vdash e \Leftarrow T$ takes an environment, an expression, and an expected type, and validates that the expression is compatible. The type environment $\Gamma$ maps variable names to pairs of type and mutability, and also records type declarations.

Because the type checker modifies the environment imperatively (each binding extends $\Gamma$ for subsequent expressions in a sequence), we write $\Gamma \vdash e \Rightarrow T \dashv \Gamma'$ when a rule extends the environment, and $\Gamma \vdash e \Rightarrow T$ when it does not. In practice, the sequencing rule threads the environment left-to-right through a sequence of expressions.

\paragraph{Synthesis rules.} Figure~\ref{fig:typing-synth} presents the synthesis rules.

Literals synthesize their corresponding base types (\textsc{Syn-String}, \textsc{Syn-Size}, \textsc{Syn-Bool}). Variable references look up the type in the environment (\textsc{Syn-Var}).

The binding rules (\textsc{Syn-Bind-Ann} and \textsc{Syn-Bind-Infer}) formalize the interaction between type annotations and the oracle rule. When a type annotation is present and the value expression is not an oracle, the value is checked against the resolved annotation (\textsc{Syn-Bind-Ann}, first premise). When the value is an oracle (a \texttt{tool} or \texttt{js} expression), the checking premise is skipped entirely, and the annotation is trusted as the type of the binding. This captures the fact that LLM outputs cannot be statically verified and the developer's annotation serves as the sole source of type information. When no annotation is present (\textsc{Syn-Bind-Infer}), the type is synthesized from the value expression. Since \texttt{tool} and \texttt{js} expressions cannot synthesize a type without an annotation (rules \textsc{Syn-Tool} and \textsc{Syn-Js}), the lack of an annotation on these expressions is a compile-time error.

\texttt{user} and \texttt{agent} expressions always synthesize \texttt{String} (\textsc{Syn-User}). \texttt{tool} and \texttt{js} expressions fail in synthesis mode (\textsc{Syn-Tool}, \textsc{Syn-Js}), requiring an annotation at the binding site to proceed.

Assignment (\textsc{Syn-Assign}) verifies that the target variable is mutable and checks the new value against the variable's declared type. Sequencing (\textsc{Syn-Seq}) threads the environment left-to-right and returns the type of the last expression.

Conditionals (\textsc{Syn-If}) check the condition against \texttt{Bool}, synthesize the then-branch, and check the else-branch against the then-branch type. When no else-branch is present, the result type is $()$. Loops (\textsc{Syn-While}) check the condition against \texttt{Bool} and return $()$. Iteration (\textsc{Syn-For}) requires the collection to have type \texttt{Schema} (since array types are represented as \texttt{Schema} at compile time), binds the iteration variable as \texttt{Schema}, and returns \texttt{Schema}.

Workflow invocation (\textsc{Syn-Eval}) synthesizes the type of the function expression. If the type is an arrow $(A \to R)$, the argument is checked against $A$ and the result type is $R$. If the type is \texttt{Schema}, the result is \texttt{Schema}, deferring the check to runtime.

Comparison operators (\textsc{Syn-Cmp}) synthesize the left operand, check the right operand against the left type, and return \texttt{Bool}.

\begin{figure}[t]
\begin{mathpar}
\inferrule*[right=Syn-String]{ }{\Gamma \vdash s \Rightarrow \texttt{String}}
\and
\inferrule*[right=Syn-Size]{ }{\Gamma \vdash n \Rightarrow \texttt{Size}}
\and
\inferrule*[right=Syn-Bool]{ }{\Gamma \vdash b \Rightarrow \texttt{Bool}}
\and
\inferrule*[right=Syn-Var]{\Gamma(x) = (T, \_)}{\Gamma \vdash x \Rightarrow T}
\and
\inferrule*[right=Syn-Bind-Ann]{\mathit{resolve}(\Gamma, T) = T' \\ \neg\mathit{oracle}(e) \Rightarrow \Gamma \vdash e \Leftarrow T'}{\Gamma \vdash \texttt{let}\; x {:} T = e \Rightarrow () \dashv \Gamma, x{:}(T', m)}
\and
\inferrule*[right=Syn-Bind-Infer]{\Gamma \vdash e \Rightarrow T}{\Gamma \vdash \texttt{let}\; x = e \Rightarrow () \dashv \Gamma, x{:}(T, m)}
\and
\inferrule*[right=Syn-User]{ }{\Gamma \vdash \texttt{user}\; s \Rightarrow \texttt{String}}
\and
\inferrule*[right=Syn-Tool]{ }{\Gamma \vdash \texttt{tool}\; s \Rightarrow \mathbf{error}}
\and
\inferrule*[right=Syn-Js]{ }{\Gamma \vdash \texttt{js}\; c \Rightarrow \mathbf{error}}
\and
\inferrule*[right=Syn-Assign]{\Gamma(x) = (T, \textit{mut}) \\ \Gamma \vdash e \Leftarrow T}{\Gamma \vdash x = e \Rightarrow ()}
\and
\inferrule*[right=Syn-Seq]{\Gamma \vdash e_1 \Rightarrow T_1 \dashv \Gamma_1 \\ \Gamma_1 \vdash e_2 \Rightarrow T_2}{\Gamma \vdash e_1;\; e_2 \Rightarrow T_2}
\and
\inferrule*[right=Syn-If]{\Gamma \vdash e_c \Leftarrow \texttt{Bool} \\ \Gamma \vdash e_t \Rightarrow T \\ \Gamma \vdash e_f \Leftarrow T}{\Gamma \vdash \texttt{if}\;(e_c)\;\{e_t\}\;\texttt{else}\;\{e_f\} \Rightarrow T}
\and
\inferrule*[right=Syn-While]{\Gamma \vdash e_c \Leftarrow \texttt{Bool} \\ \Gamma \vdash e_b \Rightarrow \_}{\Gamma \vdash \texttt{while}\;(e_c)\;\{e_b\} \Rightarrow ()}
\and
\inferrule*[right=Syn-For]{\Gamma \vdash e_c \Rightarrow \texttt{Schema}}{\Gamma, x{:}(\texttt{Schema}, \textit{imm}) \vdash \texttt{for}\; x\; \texttt{in}\; e_c \;\{e_b\} \Rightarrow \texttt{Schema}}
\and
\inferrule*[right=Syn-Eval]{\Gamma \vdash e_f \Rightarrow (A \to R) \\ \Gamma \vdash e_a \Leftarrow A}{\Gamma \vdash \texttt{eval}\; e_f\;(e_a) \Rightarrow R}
\and
\inferrule*[right=Syn-Cmp]{\Gamma \vdash e_1 \Rightarrow T \\ \Gamma \vdash e_2 \Leftarrow T}{\Gamma \vdash e_1 \;\mathit{op}\; e_2 \Rightarrow \texttt{Bool}}
\end{mathpar}
\caption{Synthesis rules. The notation $\Gamma \vdash e \Rightarrow T \dashv \Gamma'$ indicates that the rule extends the environment. In \textsc{Syn-Bind-Ann}, the checking premise is applied only when $\neg\mathit{oracle}(e)$ holds. When $\mathit{oracle}(e)$ is true (\texttt{tool} or \texttt{js} expressions), the annotation type is trusted without verification. The mutability marker $m$ is \textit{imm} for \texttt{let} and \textit{mut} for \texttt{var}.}
\label{fig:typing-synth}
\end{figure}

\paragraph{Checking rules.} Figure~\ref{fig:typing-check} presents the checking rules. The oracle rule (\textsc{Chk-Oracle}) is the most distinctive. When checking a \texttt{tool} or \texttt{js} expression against an expected type, the check succeeds unconditionally. This formalizes the treatment of LLM tool calls as type-theoretic oracles whose output type is determined by the developer's annotation rather than by static analysis of the expression.

Sequence checking (\textsc{Chk-Seq}) synthesizes all expressions except the last, then checks the last against the expected type. This allows type information to flow backward through sequences. Conditional checking (\textsc{Chk-If}) propagates the expected type into both branches, allowing both to be checked against the same target type. Loop and iteration checking (\textsc{Chk-While}, \textsc{Chk-For}) accept any expected type, since these expressions are evaluated for side effects.

The subsumption rule (\textsc{Chk-Sub}) is the bridge between the two modes. When no other checking rule applies, the type checker synthesizes a type for the expression and verifies that the synthesized type is a subtype of the expected type.

\begin{figure}[t]
\begin{mathpar}
\inferrule*[right=Chk-Oracle]{\mathit{oracle}(e)}{\Gamma \vdash e \Leftarrow T}
\and
\inferrule*[right=Chk-Seq]{\Gamma \vdash e_1 \Rightarrow \_ \dashv \Gamma_1 \\ \Gamma_1 \vdash e_2 \Leftarrow T}{\Gamma \vdash e_1;\; e_2 \Leftarrow T}
\and
\inferrule*[right=Chk-If]{\Gamma \vdash e_c \Leftarrow \texttt{Bool} \\ \Gamma \vdash e_t \Leftarrow T \\ \Gamma \vdash e_f \Leftarrow T}{\Gamma \vdash \texttt{if}\;(e_c)\;\{e_t\}\;\texttt{else}\;\{e_f\} \Leftarrow T}
\and
\inferrule*[right=Chk-While]{\Gamma \vdash e_c \Leftarrow \texttt{Bool} \\ \Gamma \vdash e_b \Rightarrow \_}{\Gamma \vdash \texttt{while}\;(e_c)\;\{e_b\} \Leftarrow T}
\and
\inferrule*[right=Chk-For]{\Gamma \vdash e_c \Rightarrow \texttt{Schema} \\ \Gamma, x{:}(\texttt{Schema}, \textit{imm}) \vdash e_b \Rightarrow \_}{\Gamma \vdash \texttt{for}\; x\; \texttt{in}\; e_c \;\{e_b\} \Leftarrow T}
\and
\inferrule*[right=Chk-Sub]{\Gamma \vdash e \Rightarrow S \\ S \leq T}{\Gamma \vdash e \Leftarrow T}
\end{mathpar}
\caption{Checking rules. \textsc{Chk-Oracle} accepts any expected type for \texttt{tool} and \texttt{js} expressions. \textsc{Chk-Sub} bridges synthesis and checking via subtyping.}
\label{fig:typing-check}
\end{figure}

\paragraph{Program checking.} The type checker processes an entire program in two passes. In the first pass, it collects all declarations into the environment. For each \texttt{type} declaration $\texttt{type}\; X = T$, the type $T$ is registered under the name $X$. For each workflow declaration $\texttt{workflow}\; f(\overline{x_i{:}T_i}) \to T\; \{e\}$, a curried arrow type is constructed by folding the parameter types right-to-left:

\[
\mathit{wftype}(x_1{:}T_1, \ldots, x_k{:}T_k, T) = T_1 \to (T_2 \to (\cdots \to (T_k \to T)))
\]

This curried type is bound to the workflow name $f$ in the environment, making workflows available as typed values for use in \texttt{eval} expressions.

In the second pass, the type checker verifies each workflow body. For a workflow $\texttt{workflow}\; f(\overline{x_i{:}T_i}) \to T\;\{e\}$, the checker resolves the return type $T$ in the current environment, binds the parameters $x_i{:}T_i$ as immutable, and checks the body against the resolved return type. Formally, $\Gamma, \overline{x_i{:}(T_i, \textit{imm})} \vdash e \Leftarrow T$. If the check fails, a type error with the conflicting types and their source locations is reported.


\subsection{Staged Gradual Type Checking}
\label{sec:design-staged}

The subtyping rules \textsc{S-Schema-L} and \textsc{S-Schema-R} allow \texttt{Schema}-typed values to pass through the compile-time type checker without constraint. This is intentional, since the concrete type of a schema-typed variable is determined by an LLM at runtime and does not exist at compile time. The staged gradual type checking mechanism bridges this gap by re-invoking the type checker at runtime whenever a schema value materializes.

\paragraph{The schema-to-type mapping.} When an LLM tool call produces a JSON Schema value at runtime, the interpreter converts it to a HADL type using the mapping $\sigma$ defined in Figure~\ref{fig:schema-to-type}. Primitive JSON Schema types map to their HADL counterparts. Compound types (arrays, objects, and recursive references) map to \texttt{Schema}, since the compile-time type language does not include record or array types. This means that the staged re-check validates the overall type structure of downstream expressions but does not statically verify field-level access on objects. Field-level property access is performed by the embedded JavaScript engine at runtime, operating on the underlying JSON values.

\begin{figure}[t]
\[
\sigma(\tau) = \begin{cases}
() & \text{if } \tau = \texttt{null} \\
\texttt{Bool} & \text{if } \tau = \texttt{boolean} \\
\texttt{Size} & \text{if } \tau = \texttt{integer} \text{ or } \tau = \texttt{number} \\
\texttt{String} & \text{if } \tau = \texttt{string} \text{ or } \tau = \texttt{enum} \\
\texttt{Schema} & \text{if } \tau = \texttt{array} \text{ or } \tau = \texttt{object} \text{ or } \tau = \texttt{\$ref}
\end{cases}
\]
\caption{The schema-to-type mapping $\sigma$. Converts a runtime JSON Schema type $\tau$ into a compile-time HADL type. Primitive schemas map to concrete HADL types. Compound schemas remain as \texttt{Schema}.}
\label{fig:schema-to-type}
\end{figure}

\paragraph{Runtime re-checking algorithm.} The staged re-check is triggered during the evaluation of expression sequences. When the interpreter evaluates a binding of the form $\texttt{let}\; x {:} \texttt{Schema} = e$ and the tool call $e$ produces a JSON Schema value, the following steps execute.

\begin{enumerate}
\item The interpreter deserializes the LLM's response as a JSON Schema value.
\item The mapping $\sigma$ converts the schema to a concrete HADL type $T$.
\item The type $T$ is registered in the environment as the concrete type for variable $x$.
\item A new type environment $\Gamma'$ is constructed from the current runtime state. For each variable in scope, its type is determined from its runtime value (or from its registered schema type if one exists). Resolved schema types are also registered as type declarations, so that $x$ can be used in type annotation positions in subsequent bindings.
\item The type checker is re-invoked in synthesis mode over the remaining expressions in the sequence, using $\Gamma'$.
\end{enumerate}

When a schema-typed variable is mutated via assignment, the interpreter re-parses the new value as a JSON Schema, updates the registered concrete type via $\sigma$, and triggers an authoritative re-check of the remaining expressions. If this re-check fails and the \texttt{@retries} pragma is active, the self-healing mechanism identifies the causal tool call through provenance tracking and re-invokes it with the type error as feedback. We describe the self-healing mechanism and provenance tracking in detail in Section~\ref{sec:implementation}.

\paragraph{Guarantees and limitations.} The staged mechanism provides two guarantees. First, at compile time, all statically known types are verified before any LLM call is made, catching undefined variables, type mismatches, and missing annotations. Second, at runtime, the type checker is re-invoked with the concrete schema type substituted into the environment, verifying that downstream bindings and subtyping relationships are consistent with the materialized type. The main limitation is that compound JSON Schema types (objects and arrays) map to \texttt{Schema} under $\sigma$, so the re-check does not verify field-level structure at the HADL type level. Field access on schema-typed objects is performed by the JavaScript engine at runtime, and structural validation of JSON values against their schemas is handled by the runtime schema system described in Section~\ref{sec:implementation}.
