\section{Evaluation}
\label{sec:evaluation}

We evaluate \pact{} on the clinical-trial workflow of
Figure~\ref{fig:clinical-trial}. The evaluation is a focused case study
of gradual schema typing and per-invocation Cedar authorization. We
compare \pact{} against baselines that must implement the same
properties manually, and we measure whether early causal repair avoids
repeated downstream failures.

We compare five systems, namely \pact{}, plain LangGraph, LangGraph
with assertions, plain DSPy, and DSPy with assertions. Each system is
run $K=3$ times on the same clinical-trial task, using a fixed scripted
transcript and the same LLM backend. Each baseline receives the same
prompt text as \pact{}, including the schema \texttt{gen} prompt, the
policy \texttt{gen} prompt, the patient-ID prompt, and the same bare
visit prompt for each generated patient visit. In particular, the visit
prompt does not inject the generated schema. This setup isolates the
effect of the workflow language. Either the system tracks and enforces
the schema it generated, or it must recover manually after a bad
downstream value has already been produced. 

\begin{sloppypar}
The two plain baselines access \texttt{visit["cost"]} and
\texttt{visit["patient\_id"]} directly, with no validation
and no authorization check at the call site. The assertion baselines add
Pydantic field-presence checks and a developer-written
\texttt{check\_authorization()} call at each LLM call site. When a field
check fails, the assertion baseline retries the failed visit generation
with the schema and error message manually injected into the prompt.
\end{sloppypar}

\subsection{Results}
\label{sec:eval-results}

\input{figures/eval-table}

\paragraph{Plain baselines fail late.}
LangGraph and DSPy without assertions fail on every run. The bare visit
prompt \texttt{"Generate a visit for pid in trial"} produces a medically
plausible CRF record with fields such as
\texttt{tender\_joint\_count} and \texttt{das28\_crp}, but without a
\texttt{cost} field. This happens because no constraint communicated to
the LLM that \texttt{cost} would be accessed later. The direct field
access \texttt{visit["cost"]} raises a \texttt{KeyError}, so the
workflow crashes only after the invalid visit has already been
generated. The three missed violations per system reflect that these
frameworks provide no language-level mechanism to connect the generated
schema with later field accesses.

\paragraph{\pact{} repairs the causal artifact.}
\pact{} completed all three runs. In each run, the initial
\texttt{crf} schema generation omitted the \texttt{cost} field, a
tracking field not implied by the bare clinical prompt. Before any visit
data was generated, \pact{} scanned the continuation of the schema
binding, found that \texttt{visit.cost} is referenced in the loop body,
and detected that the generated schema did not provide that field. The
runtime then retried the schema-producing \texttt{gen} with the type
error as feedback. The repaired schema included \texttt{cost}, and all
subsequent visit generations conformed on the first attempt.

This is the key difference from downstream assertion checking. \pact{}
repairs the schema once at the source of the defect. It does not wait
for each visit generation to fail independently. The self-healing event
is therefore counted once per run, giving \texttt{caught}$=3$ and
\texttt{missed}$=0$ across the three runs.

\paragraph{\pact{} enforces authorization at each invocation.}
The generated policy is installed with \texttt{enforce}. After
installation, each \texttt{gen} and \texttt{agent} call is checked
against Cedar under the principal carried by the call. In the successful
\pact{} runs, the coordinator principal is permitted to generate visit
data under the installed blinding policy, and the final reporting
agent runs under the default principal. No authorization violation
occurred in these runs. The
important point is that the check is inserted by the language at every
LLM invocation. In the assertion baselines, the same property depends on
a developer-written \texttt{check\_authorization()} call at each site.

To make this concrete, consider what happens when a \texttt{gen} 
call is attempted under a principal that the installed policy does 
not permit. The Gen-PolHeal rule fires before the LLM is called 
at all. Cedar evaluates the request under the current principal, 
produces a denial, and the runtime appends that denial as a 
feedback event to the heal context. The generator is then retried 
with the Cedar denial reason included in the prompt, giving the 
LLM the opportunity to produce a value under a permitted 
principal. In the clinical-trial workflow, the
\texttt{coordinator} principal on line~6 of
Figure~\ref{fig:clinical-trial} is used to scope the visit
generation on line~13 via an explicit \texttt{as} clause; calls
without such a clause run under the default principal. A developer
who adds a new \texttt{gen} call to
the workflow without specifying an explicit \texttt{as} clause 
runs under the default principal, which is checked against the 
installed policy before the LLM is ever contacted. The policy 
violation surfaces immediately and deterministically, regardless 
of what the LLM would have produced. In the assertion baselines, 
the same guarantee depends on the developer remembering to add a 
\texttt{check\_authorization()} call at the new site. Omitting it 
produces no error and no warning. The workflow silently bypasses 
the intended policy.

\paragraph{Assertion baselines succeed with manual recovery.}
LangGraph with assertions and DSPy with assertions complete all runs,
but they need 10 LLM calls per run instead of 7. The additional calls
are caused by late repair. Each of the three visit generations first
omits \texttt{cost}, then fails the Pydantic check, then is retried with
the schema and error message manually injected into the prompt. Across
three runs, this gives \texttt{caught}$=9$ violations for the assertion
baselines, compared with \texttt{caught}$=3$ for \pact{}. The higher
count is not better behavior. It means the same schema defect is
rediscovered separately at each dependent call.

The extra retries more than double the runtime, from about 38 seconds
for \pact{} to about 81 seconds for the assertion baselines, and
increase output tokens from 1{,}732 to roughly 4{,}600. The assertion
baselines also require substantially more code. LangGraph with
assertions is 154 lines and DSPy with assertions is 135 lines, compared
with 19 lines for the \pact{} workflow.

\subsection{Threats to Validity}
\label{sec:eval-threats}

This evaluation is a focused case study rather than a broad benchmark.
It uses one workflow, one model family, and three runs per system.
Schema healing triggered on every \pact{} run because the bare clinical
prompt consistently omitted \texttt{cost}, and \pact{} repaired the
schema on the first retry. The claim is therefore not that the LLM will
always repair every schema after one retry. The claim is that, when a
structural defect is repairable, \pact{} localizes the repair to the
causal generated artifact before dependent calls execute.

The authorization comparison is partly qualitative. The assertion
baselines can be made correct by careful developers, but the property is
not enforced by the framework. Omitting a manual authorization check at
one call site silently bypasses the intended policy. In \pact{}, the
Cedar check is part of the semantics of \texttt{gen} and \texttt{agent}
after policy installation.

The authorization evaluation does not include a run in which a 
genuine Cedar denial is triggered, because doing so would require 
deliberately misconfiguring the principal structure in a way that 
falls outside the scope of this case study. The formal coverage of that case is provided by the Gen-PolHeal 
and Agent-PolHeal rules in Section~\ref{sec:sem-reduction} and 
their mechanized Lean counterparts.

The semantics defines five possible outcomes for any terminating run:
\begin{itemize}
  \item \emph{success} (reduction reaches a value configuration),
  \item \emph{type rejection} (a non-healable \texttt{gen} produced an ill-typed value and fired \rulename{Gen-TypeErr}, or the budget exhausted and fired \rulename{Gen-Budget}),
  \item \emph{policy rejection} (\rulename{Gen-PolHeal} consumed the full budget),
  \item \emph{policy-install failure} (the deny-all guard or action-type check rejected an \texttt{enforce} policy after $\MRetry$ retries), or
  \item \emph{oracle stall} (every oracle response inside budget was ill-typed at a healable $\tau$, so no rule fired and the trace got stuck).
\end{itemize}