\section{Implementation}
\label{sec:implementation}

This section describes the compilation pipeline, runtime execution model, and authorization integration of the HADL implementation. The system is implemented in Rust as a workspace of six crates excluding the vendored Copilot SDK dependency.

\subsection{Compilation Pipeline}
\label{sec:impl-pipeline}

HADL programs are processed in three phases, with no intermediate representation or code generation step between them.

\paragraph{Phase 1 -- Parsing.} The parser is generated from a PEG grammar of 52 rules using the \texttt{pest} parser generator. The grammar defines whitespace as spaces and tabs (not newlines), treats newlines as expression separators within sequences, and reserves 21 keywords. The parser transforms source text into a typed AST whose nodes correspond directly to the \texttt{Expr}, \texttt{Type}, and \texttt{Declaration} types defined in Section~\ref{sec:design-syntax}. String literals are further processed to extract backtick-delimited interpolation segments, which are stored as \texttt{StringPart::JsExpr} nodes for evaluation by the JavaScript engine at runtime.

\paragraph{Phase 2 -- Type checking.} The type checker implements the two-pass algorithm described in Section~\ref{sec:design-typing}. The first pass collects all workflow and type declarations, building curried arrow types for workflows and registering type aliases. The second pass checks each workflow body against its declared return type using the synthesis and checking judgments. All type errors carry source spans for diagnostic reporting. The type checker is also invoked at runtime during staged gradual type checking, as described in Section~\ref{sec:design-staged}.

\paragraph{Phase 3 -- Interpretation.} The interpreter walks the AST directly, evaluating each expression node recursively. There is no compilation to bytecode or intermediate representation. The interpreter is generic over an \texttt{InputProvider} trait that abstracts human interaction, allowing the same interpreter to run with standard input in headless mode or with an interactive TUI dashboard. The interpreter communicates with the dashboard through an unbounded channel of \texttt{RuntimeEvent} values that carry source spans, tool descriptions, streaming LLM output deltas, and completion status.

\subsection{Runtime Execution Model}
\label{sec:impl-runtime}

The interpreter maintains five pieces of state that together support gradual schema typing, provenance tracking, and policy enforcement.

\paragraph{Unified environment.} The variable environment stores each binding as a \texttt{Variable} struct containing four fields: the runtime JSON value, a mutability flag, an optional concrete schema type (for gradual typing), and optional provenance information (for self-healing). This unified representation prevents desynchronization between the value, its type, and its causal history. The environment also stores program-level type declarations separately, since these are workflow-scoped rather than per-variable.

\paragraph{JavaScript engine.} HADL delegates all arithmetic and property access to an embedded JavaScript engine (Boa). The engine maintains a persistent context across evaluations within a single workflow execution, so variables defined in one \texttt{js} expression remain accessible in subsequent ones. Before each evaluation, the interpreter injects all HADL variables into the JavaScript global scope by converting \texttt{serde\_json::Value} instances to Boa \texttt{JsValue} instances, supporting null, boolean, number, string, array, and object types. After evaluation, the interpreter reads back all mutable HADL variables from the JavaScript context, so that assignments like \texttt{js x = x + 1} are reflected in the HADL environment. Immutable variables are not written back. The JavaScript engine is also used to evaluate backtick expressions inside string literals. When the parser encounters a backtick-delimited segment such as \texttt{`person.name`}, it stores it as a \texttt{StringPart::JsExpr} node. At prompt evaluation time, the interpreter evaluates each such segment through the Boa engine with the current environment injected, then concatenates the results with the literal segments to produce the final string.

\paragraph{Expression evaluation.} The \texttt{eval} method is a recursive async function that pattern-matches on all expression variants. Literals return their corresponding JSON values. Variable references look up the unified environment. \texttt{let} and \texttt{var} bindings evaluate the value expression, record provenance if the value is a \texttt{tool} or \texttt{agent} call, detect \texttt{Schema} annotations to trigger gradual typing registration, and bind the result in the environment. Assignment checks mutability and, for schema-typed variables, re-parses the new value as a JSON Schema and sets the \texttt{schema\_dirty} flag to trigger re-checking. Sequencing evaluates expressions left-to-right, threading the environment, and after each expression checks whether staged re-checking should be triggered. Control flow expressions (\texttt{if}, \texttt{while}, \texttt{for}) evaluate their conditions and bodies in the standard imperative manner. The \texttt{for} expression evaluates the collection, asserts it is a JSON array, binds each element to the iteration variable, evaluates the body for each element, and collects the results into a new JSON array. Comparison operators delegate to a helper that coerces operands to floating-point numbers via \texttt{as\_f64()} for ordered comparisons, while equality and inequality use direct JSON value comparison.

\paragraph{Staged re-checking.} The sequence evaluator is the site where staged gradual type checking executes at runtime. After evaluating each expression in a sequence, the interpreter checks two conditions. First, if the expression was a \texttt{let} binding with a \texttt{Schema} type annotation and the environment now holds a resolved schema type for that variable, the interpreter performs a best-effort re-check of the remaining expressions. This re-check constructs a \texttt{TypeEnv} from the current runtime state by inferring types from runtime JSON values for ordinary variables and registering resolved schema types as type declarations for schema-typed variables. The type checker is then invoked in synthesis mode over the remaining expressions. Errors from this initial re-check are intentionally discarded, since subsequent expressions may mutate the schema before it is used. Second, if the \texttt{schema\_dirty} flag is set (indicating that a schema-typed variable was reassigned), the interpreter performs an authoritative re-check. If this re-check fails and the \texttt{@retries} pragma is active, the self-healing mechanism is invoked.

\paragraph{Self-healing via provenance tracking.} Each tool and agent call result is annotated with provenance information recording the original prompt text and the set of variable names referenced via \texttt{\{var\}} interpolation in that prompt. This information is stored in the unified environment alongside the variable's value. When the authoritative re-check fails, the interpreter walks the causal chain of the failing variable by following provenance dependencies transitively. For each variable in the chain that has provenance, the interpreter constructs a retry prompt by appending the type error message to the original prompt and re-invokes the tool call through the Copilot SDK. After each retry, the interpreter updates the variable's value and schema type in the environment and re-runs the type checker over the remaining expressions. If the re-check succeeds, execution resumes normally. If all retries are exhausted across the entire causal chain, the interpreter raises a schema type error with the full diagnostic.

\paragraph{Workflow invocation (\texttt{eval}).} The \texttt{eval} expression enables the metacircular pattern described in Section~\ref{sec:related}. At runtime, the interpreter evaluates the function expression to obtain a JSON value, deserializes it as a \texttt{Program} AST using \texttt{serde\_json::from\_value}, locates the first workflow declaration within the deserialized program, validates that the number of supplied arguments matches the workflow's parameter count, evaluates each argument expression, binds the results to the workflow's parameter names in the current environment, and evaluates the workflow body. The body executes in the current interpreter context with access to all variables in scope, providing closure-like semantics. At the type level, as described in Section~\ref{sec:design-typing}, the type checker resolves the function expression's type as an arrow type and checks the argument against the parameter type, or defers to gradual typing when the function has type \texttt{Schema}.

\paragraph{LLM integration.} The interpreter communicates with LLMs through the GitHub Copilot SDK, wrapped in a \texttt{CopilotClient} that maintains two persistent sessions. The tool session is configured with a system prompt that instructs the LLM to return only raw result values as valid JSON, with no markdown or explanation. The agent session uses a more flexible system prompt that allows open-ended reasoning while still requesting raw output. Both sessions support two execution modes. In headless mode, the client sends a prompt and collects the complete response in a single call. In streaming mode, the client subscribes to a session event stream and emits \texttt{RuntimeEvent::ReasoningDelta} and \texttt{RuntimeEvent::AssistantDelta} events for each incremental output fragment, enabling the TUI dashboard to display reasoning in real time. LLM responses are parsed by first attempting JSON deserialization, then attempting extraction from markdown code blocks, and finally falling back to a plain string value.

\subsection{Authorization Integration}
\label{sec:impl-auth}

The policy system translates HADL authorization specifications into Cedar policies and evaluates them at each tool and agent invocation.

\paragraph{Policy specification.} Policies are defined in JSON files that specify an array of permit and forbid rules, each consisting of a principal pattern, an action pattern, a resource pattern, and optional conditions. Principal patterns can match any principal, a specific node type (such as \texttt{"analyst"} or \texttt{"auditor"}), a specific node ID, or any of a set of patterns. Action patterns can match any action or specific action types (\texttt{ExecuteTool}, \texttt{ReadFile}, \texttt{WriteFile}, \texttt{InvokeAgent}). Resource patterns match any resource or any of a set of patterns. Conditions support glob-based pattern matching on the tool description via \texttt{file\_matches}, with boolean combinators \texttt{and}, \texttt{or}, and \texttt{not}. The specification also includes an enforcement configuration with a default-deny policy.

\paragraph{Cedar translation.} At startup, the runtime loads the policy JSON and translates each rule into Cedar syntax. Each HADL policy becomes a Cedar policy statement with an \texttt{@id} annotation, an effect (\texttt{permit} or \texttt{forbid}), and scope clauses for principal, action, and resource. Principal node-type constraints are translated into Cedar \texttt{when} clauses that check the \texttt{node\_type} attribute on the principal entity. Conditions are translated recursively into Cedar expressions, with \texttt{file\_matches} becoming Cedar \texttt{like} operators on the resource's \texttt{command} attribute. The generated Cedar text is parsed into a \texttt{PolicySet} using the Cedar policy engine.

\paragraph{Runtime enforcement.} Before each \texttt{tool} and \texttt{agent} invocation, the interpreter's \texttt{check\_policy} method constructs an authorization triplet. The principal is built from a monotonically increasing call-site counter and the principal name, which comes from the \texttt{as} clause if present or defaults to \texttt{"tool"} or \texttt{"agent"}. The action type is determined by the expression kind, with \texttt{agent} expressions and principals whose names end in \texttt{\_agent} producing an \texttt{InvokeAgent} action and all others producing an \texttt{ExecuteTool} action. The resource is a \texttt{NodeOutput} identified by the call-site counter. The interpreter builds Cedar entities for the principal (with a \texttt{node\_type} attribute) and the resource (with a \texttt{node\_id} attribute and, for \texttt{ExecuteTool} actions, a \texttt{command} attribute containing the tool description). These entities and the authorization request are passed to the Cedar \texttt{Authorizer}, which evaluates the compiled policy set and returns a permit or deny decision. A deny decision causes the interpreter to emit a \texttt{RuntimeEvent::Error} event and raise a \texttt{PolicyViolation} error that halts the workflow before the LLM call is made. If no policy validator is configured, the check is a no-op and the tool call proceeds unconditionally.
