\section{Introduction}
\label{sec:intro}

Large language models have moved beyond single-turn question answering into the role of autonomous agents that decompose complex tasks, invoke external tools, incorporate human feedback, and iteratively refine their outputs over extended interactions~\cite{wei2022chain,yao2022react,schick2023toolformer}. Recent work has demonstrated that these agentic capabilities can be composed into sophisticated multi-turn workflows, including hierarchical task decomposition~\cite{yao2023tree}, self-reflective refinement loops~\cite{shinn2024reflexion}, and coordinated multi-agent systems~\cite{wu2024autogen,hong2023metagpt}. Surveys of the rapidly expanding landscape~\cite{chen2024survey,huang2024understanding} reveal a common structural pattern. At their core, all of these systems execute structured \emph{plans} that branch on conditions, iterate over collections, accumulate mutable state, invoke external services, and compose partial results into final outputs. As the complexity and scale of these plans grows, from simple tool-calling loops to nested pipelines spanning hundreds of agent interactions, the need for the same engineering discipline we apply to programs becomes pressing, namely type safety, static analysis, compositional design, and principled error reporting.

Today, developers express agentic plans using general-purpose orchestration frameworks such as LangGraph~\cite{langgraph2024} and LangChain~\cite{langchain2022}, or through ad-hoc scripts in Python. These approaches treat plans as \emph{configurations of library objects} such as graphs to wire, chains to compose, and callbacks to register, rather than as programs in their own right. Four consequences follow. First, tool calls return untyped data, so mismatches between the structure a developer expects from an LLM and the structure actually produced are caught only at runtime, often after expensive API calls have already been made. Second, control flow is expressed through graph construction or callback patterns rather than through natural imperative constructs such as loops, conditionals, and variable bindings, forcing developers to reason about state transitions in a graph rather than about the sequential logic of their plan. Third, structural errors in the workflow, including undefined variables, type mismatches, and incompatible data shapes flowing between stages, surface during execution rather than before it, defeating the purpose of having a structured workflow in the first place. Fourth, the same absence of formal structure that renders type errors invisible also renders authorization intractable at the granularity that agentic systems require. Agent systems today operate with coarse-grained access control, where an agent either has access to all available tools and data, or to none. There is no mechanism to enforce fine-grained policies \emph{per invocation}, specifying which data a particular tool call may consume, whether a value produced by one agent may flow to another, or whether human approval is required before a sensitive operation proceeds. Because existing frameworks have no representation of data flow or type structure, they cannot express, let alone enforce, such policies. Authorization is treated as an external concern, bolted on after the workflow is designed.

Existing systems address parts of this problem but leave the core gap open. LangGraph~\cite{langgraph2024} provides durable execution and human-in-the-loop interrupts but requires manual graph construction and offers no static type checking. DSPy~\cite{khattab2023dspy} compiles declarative language model pipelines into optimized prompting strategies but targets single-turn structured generation, not long-running stateful workflows with mutable state and unbounded iteration. LMQL~\cite{beurer2023prompting} brings programming language ideas to LLM interaction through a query language with type constraints, representing the closest precedent in the PL literature, but focuses on prompt-level scripting rather than multi-turn agent orchestration with control flow and state. A recent declarative DSL for agent workflows~\cite{daunis2025declarative} separates workflow specification from implementation but uses DAG-based configuration without a type system or imperative control flow, and cannot express the iterative refinement patterns that characterize long-running agent interactions. AutoGen~\cite{wu2024autogen} and MetaGPT~\cite{hong2023metagpt} enable multi-agent collaboration but lack formal type systems or compile-time error detection. Critically, none of these systems integrate authorization into the execution semantics. Policy enforcement, where it exists, is handled by external middleware or ad-hoc checks scattered through application code, disconnected from the workflow's type structure and data flow. To our knowledge, no existing system brings programming language theory, specifically gradual type checking, structural subtyping, and compositional semantics, to the problem of agentic workflow specification with per-invocation policy enforcement.

Our central observation is that agentic workflows are \emph{programs} in every meaningful sense. They bind variables, branch on conditions, iterate over collections, accumulate mutable state, call external services, and compose into larger units. If plans are programs, then the mature tools of programming language design, including type systems, static analysis, effect management, and authorization semantics, should apply directly. Four specific observations guide our design. First, LLM tool calls occupy a distinctive position in the type-theoretic landscape. They behave as \emph{oracles}, black boxes that can produce any type at any call site, with the developer's annotation at the binding site as the sole source of type information. Moreover, tool calls can generate not only values but also \emph{types}, as when an LLM produces a schema that governs subsequent data. These dynamically generated types cannot be checked at compile time because they do not yet exist. This leads naturally to a gradual type checking discipline~\cite{siek2006gradual,siek2015refined} in which the type checker operates in multiple stages. The first stage runs at compile time, checking all statically known types. When an LLM call produces a new type at runtime, execution pauses, the materialized type is substituted into the program, and the type checker is re-invoked over the downstream program before execution resumes. This staged discipline re-verifies downstream expressions whenever a schema value is resolved or mutated, catching type errors introduced by LLM-generated types before dependent code proceeds. When re-checking fails, a provenance-driven self-healing mechanism re-invokes the causal tool call with the type error as feedback, giving the LLM an opportunity to correct its output. Second, the data flowing through agent workflows is naturally represented as structural records with set-of-fields semantics~\cite{cardelli1984semantics,remy1989type}, where different workflow stages care about different subsets of fields and width subtyping~\cite{pierce2002types} allows values to flow between stages without explicit conversion. Third, governance in agentic systems is not a deployment concern but an \emph{execution semantics} concern. When a tool call produces a value, the system must track who produced it and enforce where it may flow. When an agent invocation is about to execute, the runtime must validate whether the current principal is authorized to perform that action on that resource. This validation must happen \emph{before every call}, as part of the language's evaluation semantics, not as an external afterthought. Finally, treating workflows as programs yields an expressiveness benefit. Because HADL code can appear as the string output of an \texttt{agent} invocation, an outer workflow can delegate sub-task planning to an LLM that generates HADL code for subsequent parsing and execution. This \emph{architect pattern} demonstrates that the language is expressive enough to support hierarchical plan generation using its existing primitives, without requiring additional orchestration mechanisms.

\begin{figure}[t]
\begin{lstlisting}[language=hadl,showstringspaces=false]
workflow register_user() -> Schema {
  let schema: Schema = tool "Generate a user
    schema with name and UID"
  let person: schema = tool "Generate a user
    matching `schema`"
  let msg: String = agent "Write a welcome
    message for `person.name`"
  person
}
\end{lstlisting}
\caption{A HADL workflow illustrating staged gradual type checking. The first \texttt{tool} call (line~2) produces a \texttt{Schema} value at runtime, for example \texttt{\{~name:~String,~UID:~Size~\}}. The runtime converts this JSON Schema to a concrete type and re-invokes the type checker over the remaining program. The second \texttt{tool} call (line~4) uses the variable \texttt{schema} in type annotation position, so its return value is bound with the concrete type derived from the generated schema. The backtick expression \texttt{person.name} (line~6) is evaluated by the embedded JavaScript engine at runtime, which performs property access on the underlying JSON value. If the LLM produces a schema that causes downstream type errors, the self-healing retry mechanism (enabled via the \texttt{@retries} pragma) re-invokes the causal tool call with the error as feedback.}
\label{fig:example}
\end{figure}

Figure~\ref{fig:example} illustrates these mechanisms in a single workflow. The \texttt{register\_user} workflow produces a user record through three interactions. The first \texttt{tool} call generates a schema, a type that does not exist at compile time. When this call returns, the runtime converts the JSON Schema value to a concrete type and re-invokes the type checker over the remaining program. The second \texttt{tool} call uses the schema variable in type annotation position, binding \texttt{person} to a value whose type is derived from the generated schema. The \texttt{agent} call uses a backtick expression \texttt{person.name}, which is evaluated by the embedded JavaScript engine performing property access on the underlying JSON value. This example demonstrates three properties of HADL in five lines. First, LLM-generated types are first-class and can appear in type annotation positions. Second, the staged type checker re-verifies downstream bindings against the materialized schema, ensuring type consistency as new types are introduced at runtime. Third, the three primitives, namely \texttt{tool} for structured data, \texttt{agent} for natural language reasoning, and \texttt{user} (not shown here but available for human-in-the-loop interaction), compose naturally within imperative control flow.

We present HADL (Hierarchical Agent Dataflow Language), a strongly typed imperative domain-specific language for composing long-running agentic workflows. HADL provides three domain-specific primitives, namely \texttt{user}, \texttt{tool}, and \texttt{agent}, that reify human interaction, LLM tool invocation, and LLM reasoning as first-class typed expressions in an imperative language with mutable state, unbounded loops, structured iteration, and nested workflow composition. The type system implements gradual type checking~\cite{siek2006gradual,siek2015refined} in which \texttt{Schema} plays the role of the gradually-typed unknown, compatible with all static types at compile time and resolved to concrete types at runtime. The runtime schema system supports structural record subtyping with width and depth subtyping and field-order independence~\cite{cardelli1984semantics,remy1989type}, validating that JSON values conform to LLM-generated schemas. At compile time, the type checker verifies all statically known types. At runtime, when an LLM call produces a new type or schema, execution pauses, the generated type is substituted into the program, and the type checker re-verifies the downstream program before execution resumes. Tool-call expressions require explicit type annotations at binding sites, formalizing the boundary between typed program logic and untyped LLM outputs. Workflows have curried function types that compose naturally and support the extended, stateful execution patterns characteristic of real-world agent deployments. A Cedar-based~\cite{cutler2024cedar} policy layer integrates authorization triplets (principal, action, resource) directly into the interpreter loop. Before every \texttt{tool} and \texttt{agent} invocation, the runtime validates the operation against a policy store, identifying the invoking principal, the requested action, and the target resource. HADL is implemented in Rust with a three-phase pipeline consisting of PEG parsing, gradual type checking, and async tree-walking interpretation over the AST, and integrates with GitHub Copilot SDK for LLM execution.

We make the following contributions, reflecting a synthesis of programming language design, type theory, and agentic systems engineering.

\begin{enumerate}
	\item We design HADL, an imperative DSL in which \texttt{user}, \texttt{tool}, and \texttt{agent} are first-class typed expressions and workflows have curried function types, providing natural imperative syntax for multi-turn agent orchestration with mutable state, unbounded loops, structured iteration, and nested composition. Compile-time checking includes diagnostic error categories and source-span reporting, catching workflow errors before any LLM call is made.
	\item We adapt gradual type checking to the agentic domain, introducing a staged typing discipline in which the type checker runs at compile time over statically known types and is re-invoked at runtime whenever an LLM call materializes a new type. Tool calls require binding-site type annotations, formalizing the boundary between typed program logic and untyped oracle outputs. The \texttt{Schema} type acts as the gradually-typed unknown, compatible with all static types, and is resolved to concrete types at runtime. The runtime schema system implements structural record subtyping with set-of-fields semantics, supporting width subtyping, depth subtyping, and field-order independence. A provenance-driven self-healing mechanism automatically retries causal tool calls when schema re-checking fails.
	\item We integrate Cedar-based authorization into the runtime evaluation loop, enforcing (principal, action, resource) triplets at each tool and agent invocation, embedding governance directly in the language's evaluation rules rather than treating it as a deployment-time concern.
	\item We present a complete implementation in Rust comprising a PEG parser, a gradual type checker, and an async tree-walking interpreter that evaluates the AST directly. The runtime includes an embedded JavaScript engine for arithmetic and expression evaluation, and integrates with the GitHub Copilot SDK for LLM execution. We evaluate the system on case studies including code review, hierarchical planning, and iterative refinement.
\end{enumerate}

The remainder of this paper is organized as follows. Section~\ref{sec:overview} presents the language design through motivating examples. Section~\ref{sec:design} formalizes the type system. Section~\ref{sec:implementation} describes the compilation pipeline, runtime execution model, and authorization integration. Section~\ref{sec:evaluation} evaluates the system through case studies examining gradual type checking, self-healing, and policy enforcement in practice. Section~\ref{sec:related} discusses related work, and Section~\ref{sec:conclusion} concludes.