\documentclass[acmsmall,screen,review]{acmart}
\input{lib}

%% Listings setup for HADL code examples
\usepackage{listings}
\usepackage{color}
\lstdefinelanguage{hadl}{
  keywords={workflow, let, var, gen, agent, ask, say, if, else, while, for, in, type, enforce, eval, as, js},
  keywordstyle=\bfseries,
  sensitive=true,
  comment=[l]{//},
  morestring=[b]",
  stringstyle=\itshape,
}
\lstset{
  basicstyle=\ttfamily\small,
  columns=flexible,
  keepspaces=true,
  showstringspaces=false,
  numbers=left,
  numberstyle=\tiny\color{gray},
  numbersep=5pt,
  xleftmargin=2em,
  frame=single,
  framesep=3pt,
  aboveskip=1em,
  belowskip=1em,
  escapeinside={(*@}{@*)},
}

%% Bibliography style
\citestyle{acmauthoryear}

%% Remove ACM-specific metadata for submission
\settopmatter{printfolios=true,printccs=false,printacmref=true}

%% Document metadata
\acmJournal{PACMPL}
\acmVolume{0}
\acmNumber{PLDI}
\acmArticle{0}
\acmYear{2026}
\acmMonth{6}
\acmDOI{}

%% Copyright - for submission
\setcopyright{none}

\begin{document}

\title{Hierarchical Agentic Dataflow Language: Plans All the Way Down}

\authoranon{
\author{First Author}
\affiliation{%
  \institution{Institution}
  \city{City}
  \state{State}
  \country{Country}
}
\email{first@example.com}

\author{Second Author}
\affiliation{%
  \institution{Institution}
  \city{City}
  \state{State}
  \country{Country}
}
\email{second@example.com}
}

\keywords{domain-specific languages, agentic workflows, gradual typing, authorization, Cedar, LLM agents}

\begin{abstract}
  Effective use of LLM CLIs requires meticulous use of plan mode. A plan is a sequence of steps in natural language to be executed by the agent. Our central observation is that agentic plans are programs, which we call \emph{agentic workflows}. If plans are programs, then type systems, static analysis, and authorization semantics should apply directly. A deeper observation is that the LLM is not merely a value producer but a generator of program structure. Gen calls can produce types that govern subsequent data, workflows that will be executed as subroutines, and policies that restrict future operations.
  We present HADL, a typed imperative domain-specific language in which human interaction, LLM gen invocation, and LLM reasoning are first-class typed expressions composable with mutable state, unbounded loops, structured iteration, and nested workflow composition. The central mechanism is \emph{gradual validation}, under which values, types, workflows, and policies produced by an LLM are checked against their declared types at the moment they are put to use, using the same type discipline that governs the rest of the program rather than a separate runtime check bolted onto a static one. When gradual validation rejects an LLM output, HADL \emph{self-heals} by rerunning the generator with the violated constraint and the recorded error, turning type errors on LLM output into bounded retries rather than aborts.
\end{abstract}

\maketitle

\input{intro}
\input{overview}
\input{design}
\input{semantics}
\input{implementation}
% \input{evaluation}
\input{related}
\input{conclusion}

%% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\input{appendix}

\end{document}